# Regression: Creating models to predict future observations {#regression}

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(r4np)
```

Regressions are an interesting area of data analysis since it enables us to make very specific predictions about the future incorporating different variables at the same time. As the name implies, regressions 'regress', i.e. they draw on past observations to make predictions about future observations. Thus, any analysis incorporating a regression makes the implicit assumption that the future can be best explained by the past.

I once heard someone refer to regressions as driving a car by looking at the rear-view mirror. As long as the road is straight, we will be able to successfully navigate the car. However, if there is a sudden turn, we might drive into the abyss. This makes it very clear when and how regressions are can be helpful.

Regressions are also a machine learning method, which falls under models with supervised learning. If you find machine learning fascinating, you might find the book "Hands-on Machine Learning with R" [@boehmke2019hands] very insightful and interesting.

In the following chapters we will cover three common types of regressions:

-   Single linear regression

-   Multiple regression

-   Hierarchical regression

These three types will allow you to perform any other type of linear regression you could think of. We can further distinguish two approaches to modelling via regressions:

-   *hypothesis testing*: A regression model is defined ex-ante

-   *machine learning*: A model is developed based on empirical data

In the following chapters we will slightly blur the lines between both approaches.

All our regressions will be performed using the `covid` dataset of the `r4np` package to investigate whether certain factors can predict COVID numbers in different countries. I felt, this book would not have been complete without covering this topic. After all, I wrote this book during the pandemic and it likely will mark a dark chapter in human history.

## Single linear regression {#single-linear-regression}

A single linear regression looks very similar to a correlation (see Chapter \@ref(correlations), but it is different in that it defines which variable affects another variable, i.e. a single direction relationship. I used the terms dependent variable (DV) and independent variable (IV) previously when comparing groups (see Chapter \@ref(comparing-groups), and we will use them here again. In group comparisons, the independent variable was usually a `factor`, but in regressions we can use data that is not a categorical variable, i.e. `integer`, `double`, etc.

While I understand that mathematical equations can be confusing, with regressions, they are fairly simple to understand. Also, when writing our models in R, we will continuously use a `formula` to specify our regression. A single linear regression consists of one independent variable and one dependent variable:

::: {align="center"}
$$
DV = \beta_{0} + IV * \beta_{1} + error
$$
:::

Beta ($\beta$) represents the coefficient of the independent variable, i.e. how much a change in IV causes a change in DV. For example, a one unit change in the IV might mean that the DV changes by two units of IV:

::: {#single-linear-regression-example align="center"}
$$
DV = \beta_0 + 2 * IV + error
$$
:::

If we ignore $\beta_0$ and $error$ for a moment, we find that that if $IV = 3$, our $DV = 2*3 = 6$. Therefore, if $IV = 5$, we find that $DV = 10$, and so on. According to this model, DV will always be twice as large as IV.

You might be wondering what $\beta_0$ stands for. It indicates an offset for each value, also called the intercept. Thus, no matter which value we choose for IV, DV will always be $\beta_0$ different from IV. It is a constant in our model. This can be best explained by visualising a regression line. Pay particular attention to the to the expressions after `function(x)`

```{r beta zero and beta one explained, echo=TRUE}
# A: Two models with different beta(0)
ggplot() +
  geom_function(fun = function(x) 2 * x, colour = "red") +
  geom_function(fun = function(x) 1 + 2 * x, colour = "blue") +
  see::theme_modern()

# B: Two models with the same beta(0), but different beta(1)
ggplot() +
  geom_function(fun = function(x) 2 * x, colour = "red") +
  geom_function(fun = function(x) 3 * x, colour = "blue") +
  see::theme_modern()
```

Plot B also shows what happens if we change $\beta_1$, i.e. the slope. The two models both have the same intercept (and therefore the same origin), but the blue line ascends quicker than the red one, because its $\beta_1$ is higher than the one for the red model.

Lastly, the $error$ component in the regression model refers to the deviation of data from this regression lines. Ideally, we want this value to be as small as possible.

### Fitting a regression model by hand, i.e. trial and error {#fitting-a-regression-model-by-hand}

If this sounds all awfully theoretical, let's try to fit a regression model by hand. First we need to consider what our model should be able to predict. Let's say that the number of COVID-19 cases predicts the number of deaths due to COVID-19. Intuitively we would assume this should be a linear relationship, because the more cases there are, the more likely we find more deaths caused by it.

```{r Finding a regression equation by hand p1, echo=TRUE}
# We only select most recent numbers, i.e. "2021-08-26"
# and countries which have COVID cases
covid_sample <- covid %>%
  filter(date_reported == "2021-08-26" &
           cumulative_cases != 0)

covid_sample %>%
  ggplot(aes(x = cumulative_cases,
             y = cumulative_deaths)) +
  geom_point()
```

This data visualisation shows us not much. We can see that there are three countries, which appear to have considerably more cases than most other countries. Thus, all other countries are crammed together in the bottom left corner. To improve this visualisation without removing the outliers, we can rescale the x and y axis using the function `scale_x_continuous()` and `scale_y_continuous()`.

```{r Finding a regression equation by hand p2, echo=TRUE, warning=FALSE}
covid_sample %>%
  ggplot(aes(x = cumulative_cases,
             y = cumulative_deaths)) +
  geom_point() +
  scale_x_continuous(trans = "log") +
  scale_y_continuous(trans = "log")
```

As we can see, the scatterplot is now easier to read and the dots are more spread out. This reveals that there is quite a strong relationship between `cumulative_cases` and `cumulative_deaths`. However, similar to before, we should avoid outliers when performing our analysis. For the sake of simplicity, in this section I will limit the number of countries included in our analysis, which also removes the requirement of using `scale_x_continuous()` and `scale_y_continuous()`.

```{r Finding a regression equation by hand p3, echo=TRUE, warning=FALSE}
covid_sample <- covid_sample %>%
  filter(date_reported == "2021-08-26" &
           cumulative_cases >= 2500 &
           cumulative_cases <= 150000 &
           cumulative_deaths <= 3000)

plot <- covid_sample %>%
  ggplot(aes(x = cumulative_cases,
             y = cumulative_deaths)) +
  geom_point()

plot
```

Through trial and error we can try to fit a linear line on top by adjusting the beta values. This is effectively what we hope to achieve with a regression: the best $\beta$ values which best explain our data. Let's start with the basic assumption of $y = x$ without specific $\beta$s, i.e. they are zero.

```{r Finding a regression equation by hand p4, echo=TRUE, warning=FALSE}
plot +
  geom_function(fun = function(x) x, colour = "red")
```

What we try to achieve is that the red line fits nicely inside the cloud of dots. Our very simple model provides a very poor fit to our data points, because all the dots are way below it. This makes sense, because $y = x$ would imply that every COVID-19 case leads to a death, i.e. everyone with COVID did not survive. From our own experience we know that this is luckily not true. Ideally we want the line to be less steep, because our first model does not make much sense. We can do this by adding a $\beta_1$ to our equation. Maybe only 2% of people who got COVID-19 might not have recovered, i.e. $\beta_1 = 0.02$.

```{r Finding a regression equation by hand p5, echo=TRUE, warning=FALSE}
plot +
  geom_function(fun = function(x) 0.02*x, colour = "red")
```

This time the line looks much more aligned with our observations. One could argue that it might have to move a little to the right as well, to cover the observations at the bottom a bit better. Therefore, we should add a $\beta_0$ to our equation, e.g. `-50`.= to move it to the right and tweak the $\beta_1$ ever so slightly.

```{r Finding a regression equation by hand p6, echo=TRUE, warning=FALSE}
plot +
  geom_function(fun = function(x) -50 + 0.015*x, colour = "red")
```

We finished creating our regression model. If we wanted to express it as a formula we would write $DV = -5 + 0.015 * IV$. We could now use this model to predict how high COVID cases likely will be in other countries.

### Fitting a regression model computationally {#fitting-a-regression-model-computationally}

Estimating a regression model by hand is not ideal and far from accurate. Instead, we would compute the $\beta$s based on our observed data, i.e. `cumulative_cases` and `cumulative_deaths`. We can use the function `lm()` to achieve this. I also rounded the all `numeric` values to two decimal places to make the output easier to read. We also use `tidy()` to retrieve a cleaner output from the computation.

```{r Computing a regression, echo=TRUE}
# classic r
m0 <- lm(cumulative_deaths ~ cumulative_cases, data = covid_sample)
broom::tidy(m0) %>% mutate(across(where(is.numeric), round, 2))
```

We first might notice that the `p.value` indicates that the relationship between `cumulative death` and `cumulative_cases` is significant. Thus, we can conclude that countries with more COVID cases also suffer higher numbers of people who do not successfully recover from it. However, you might be wondering where our $\beta$ scores are. They are found where it says `estimate`. The standard error (`std.error`) denotes the `error` we specified in the [previous equation](#single-linear-regression). We find that in the first row we get the $\beta_0$, i.e. the one for the intercept which is 88.10. This one is larger than what we estimated, i.e. `-50`. However, $\beta_1$ is 0.01, which means we have done a very good job in guessing this estimate. Still, it becomes hopefully obvious that it is much easier to use the function `lm()` to estimate a model than 'eyeballing it'.

We can now visualise the computed model (in blue) and our guessed model (in red) in one plot and see the differences. The plot shows that we have not been too far off. However, it was relatively easy to fit a model onto the observed data in this case. Often, it is much more difficult, especially when more than two variables are involved.

```{r Showing computed and guessed regression, echo=TRUE}
plot +
  geom_function(fun = function(x) -50 + 0.015*x, colour = "red") +
    geom_function(fun = function(x) 88.1 + 0.01*x, colour = "blue")
```

With our final model computed, we also need to check its quality in terms of predictive power based on how well it can actually explain our observed data. We have tested models before when we looked at confirmatory factor analyses for latent variables (see Chapter \@ref(latent-constructs). This time we want to know how accurate our model is in explaining observed data and therefore how accurate it will be predicting future observations. The package `performance` offers a nice little shortcut to compute many different things at once:

-   `check_model()`: Checks for linearity, homogeneity, collinearity and outliers

-   `model_performance()`: Tests the quality of our model.

For now, we are mainly interested in the performance of our model. So, we can compute it the following way:

```{r Single regression model performance, echo=TRUE}
performance::model_performance(m0)
```

There are quite a number of performance indicators and here is how to read them:

-   `AIC` stands for *Akaike Information Criterion* and the lower the score the better the model.

-   `BIC` stands for *Bayesian Information Criterion* and the lower the score the better the model.

-   `R2` stands for *R squared* ($R^2$) and is also known as the coefficient of determination. It measures how much the independent variable can explain the variance in the dependent variable. In other words, the higher $R^2$ the better is our model, because more of the variance can be explained by our model. $R^2$ falls between 0-1, where 1 would imply that our model can explain 100% of the variance in our sample. $R^2$ is also considered a *goodness-of-fit measure*.

-   `R2 (adj.)` stands for *adjusted R squared*. The adjusted version of $R^2$ becomes important if we have more than one preditor (i.e. independent variable) in our regression. The adjustment of $R^2$ accounts for the number of independent variables in our model. Thus, it is possible for us to compare different models with each other, even though they might have different numbers of predictors. It is important to note that $R^2$ will always increase if we add more predictors.

-   `RMSE` stands for *Root Mean Square Error* and is an indicate how small or large the prediction error of the model is. Conceptually, it aims to measure the average deviations of values from our model when we attempt predictions. The lower the score the better, i.e. a score of 0 would imply that our model perfectly fits the data, which is likely never the case in the field of Social Sciences. The RMSE is particularly useful when trying to compare models.

-   `Sigma` stands for the standard deviation of our residuals (the difference between predicted and empirically observed values) and is therefore a measure of prediction accuracy. `Sigma` is *'a measure of the average distance each observation falls from its prediction from the model'* [@gelman2020regression, p. 168].

Many of these indices will become more relevant when we compare models. However, $R^2$ can also be meaningfully interpreted without a reference model. We know that the bigger $R^2$ the better. In our case it is `0.548` which is very good considering that our model consist of only one predictor only. It is not easy to interpret whether an particular $R^2$ value is good or bad. In our simple single linear regression, $R^2$ is literally 'r squared', which we already know from correlations and their effect sizes (see Table \@ref(tab:effect-size-cohen). Thus, if we take the square root of $R^2$ we can retrieve the correlation coefficient, i.e. $r = \sqrt{R^2} = \sqrt{0.548} = 0.740$. According to @cohen1988statistical, this would count as a large effect size.

However, for multiple regressions, the situation is slightly more complicated, but the interpretation of $R^2$ and its adjusted version remain largely the same.

Once you have a model and it is fairly accurate, you can start making predictions. This can be achieved by using our model object `m0` in combination with the function `predict()`. However, first we should define a set of values for our independent variable, i.e. `cumulative_cases`, which we store in a tibble using the `tribble()` function.

```{r Predict values from linear regression, echo=TRUE}
df_predict <- tribble(
  ~cumulative_cases,
                100,
               1000,
              10000,
             100000
)
predict(m0, newdata = df_predict)
```

As a result we find out how many likely deaths from COVID have to be expected based on our model for each value in our dataset.

Single linear regressions are simple and a good way to introduce novice scholars to modeling social phenomena. However, hardly ever will find that a single variable can explain enough variance to be a useful model. Instead, we most likely can improve the majority of our single regression models by considering more variables in the form of multiple regressions.

## Multiple regression {#multiple-regression}

Multiple regressions expand single linear regressions by allowing us to add more variables. Maybe less surprising, computing a multiple regression is similar to a single regression in R, because it requires the same function, i.e. `lm()`. However, we add more IVs. Therefore, the equation we used before needs to be modified slightly by adding more independent variables each of which will have its own $\beta$ value:

::: {align="center"}
$$
DV = \beta_{0} + IV_{1} * \beta_{1} + IV_{2} * \beta_{2} + ... + IV_{n} * \beta_{n} + error
$$
:::

In the last section we wanted to know how many people will likely not recover from COVID. However, it might be even more interesting to understand how we can predict new cases and prevent casualties from the outset. Since I live in the United Kingdom at the time of the pandemic, I am curious to know whether certain COVID measures helped to reduce the number of new cases there. To keep it more interesting, I will also add Germany to the mix since it has shown to be very effective in handling the pandemic relative to other European countries. Of course, feel free to pick different countries (maybe the one you live in?) to follow with my example. In Chapter \@ref(moderated-regressions) it will become obvious why I chose two countries (\#spoiler-alert).

First, we create a dataset that only contains information from the `United Kingdom` and `Germany`[^regressions-1], which means we use `filter()`.

[^regressions-1]: Germany is called 'Deutschland' in German, hence the abbreivation 'DEU' according to the [ISO country code](https://www.iso.org/obp/ui/#iso:code:3166:DE "ISO country code"){target="blank"}.

```{r MR filter out UK and DEU as a sample}
covid_uk_ger <- covid %>%
  filter(iso3 == "GBR" | iso3 == "DEU")
```

In a next step, we might want to know how new cases are distributed over time. It is always a good idea to inspect the dependent variable to also get a feeling of how much variance there is in your data. Having a bigger range of data values is ideal because the regression model will be able to consider low and high values of the dependent variable instead of just high or low scores. Let's plot the DV `new_cases` across time to see when and how much new COVID cases had to be reported.

```{r Multiple regression plotting new cases, echo=TRUE}
covid_uk_ger %>%
  ggplot(aes(x = date_reported,
             y = new_cases)) +
  geom_col()
```

We can tell that there are different waves of new cases. As such, we should find that there are certain variables that should help explain when `new_cases` are high and when they are low. If you have hypotheses you want to test, you would already know which variables to include in your regression. However, in our case, we do not really have a hypothesis based on our prior reading or other studies. Thus, we pick variables of interest that we suspect could help us with modelling new COVID cases. We can be fairly certain that the number of new COVID cases should be lower if there are more safety measures in place - assuming that they are effective and adhered to, of course. The `covid` dataset includes such information evaluated by the [WHO](https://covid19.who.int/info/ "WHO"){target="blank"}, i.e. `masks`, `travel`, `gatherings`, `schools` and `movements`. Remember, you can always find out what these variables stand for by typing `?covid` into the console. A higher value for these variables indicates that there were more safety measures. Scores can range from 0 (i.e. no measures) to 100 (all WHO measures taken).

We simply add these variables by using the `+` symbol in the `lm()` function.

```{r Multiple regression, echo=TRUE}
# Create our model
m1 <- lm(new_cases ~ masks + movements + gatherings + schools + businesses + travel,
         data = covid_uk_ger)

# Inspect the model specifications
broom::tidy(m1)

# Evaluate the quality of our model
performance::model_performance(m1)
```

Overall (and purely subjectively judged), the model is not particularly great, because even though we added so many variables, the $adjusted \ R^2$ is not particularly high, i.e. 'only' 0.229. As mentioned earlier, for multiple regression it is better to look at $adjusted \ R^2$, because it adjusts for the number of variables in our model and makes comparison of multiple models easier. There are a couple more important insights gained from this analysis:

-   All variables appear to be significant and therefore help to explain new cases.

-   The variables `masks`, `movements` and `gatherings` seem to reduce `new_cases`, i.e. they have negative estimates ($\beta$).

-   However, `schools`, `businesses`, `travel` have a positive effect on `new_cases`.

Especially the last point might appear confusing. How can measures taken increase the number of new COVID cases? Should we avoid them? What we have not considered in our regression is the fact that measures might be put in place to reduce the number of new cases rather than to prevent them. Thus, it might not be the case that `schools`, `businesses` and `travel` predict higher `new_cases`, but rather the opposite, i.e. due to higher `new_cases` the measures for `schools`, `businesses` and `travel` were tightened, which later on (with a time lag) led to lower `new_cases`. Thus, the relationships might be a bit more complicated, but to keep things simple, we accept the fact that with our data we face certain limitations (as is usually the case).

However, there are a couple of things we overlooked when running this regression. If you are familiar with regressions already, you might have been folding the hands over your face and bursted into tears about the blasphemeous approach to linear regression modelling. Let me course-correct at this point.

Similar to other parametric approaches to conducting research, we need to test for sources of bias, linearity, normality and homogeneity of variance (or the unpronounceable word 'homoscedasticity'). Since multiple regressions consider more than one variable we have to consider these crieria in light of other variables. As such, we have to draw on different tools to assess our data. There are certain pre- and post-tests we have to perform to fully assess and develop a multiple regression model:

-   *pre-test*: We need to consider whether there are any outliers and whether all variables are parametric.

-   *post-test*: Do our independent variables correlate very strongly with each other, i.e. are there issues of multiple collinearity.

We already covered aspects of linearity, normality and homogeneity of variance. However, outliers and collinearity have to be reconsidered for multiple regressions.

### Outliers in multiple regressions {#outliers-in-multiple-regressions}

While it should be fairly clear by now why we need to handle outliers (remember Chapter \@ref(dealing-with-outliers)), the approach we take is somewhat different when we need to consider multiple variables at once. Instead of identifying outliers for each variable independently, we have to consider the interplay of variables as well. In other words, we need to find out how an outlier in our independent variable affects the overall model rather than just on single other variable. In short, we need a different technique to assess outliers. By now, you might not be shocked to find out that there is more than one way of identifying outliers in regressions and many different ways to compute them in R. [@cohen2014applied] distinguishes between where one can find outliers in the model as summarised in Table \@ref(tab:outliers-in-multiple-regressions). I offer a selection of possible ways of computing the relevant statistics, but this list is not exhaustive. For example, many of these statistics can also be found by using `influence.measures()`.

+------------------------+--------------------------------------+---------------------------------------------------------------------------------+
| Outlier in?            | Measures                             | function in R                                                                   |
+========================+======================================+=================================================================================+
| *Dependent variable*   | -   Internally studentized residuals | -   `rstandard()` or `fortify()`                                                |
+------------------------+--------------------------------------+---------------------------------------------------------------------------------+
| *Dependent variable*   | -   Externally studentized residuals | -   `rstudent()`                                                                |
+------------------------+--------------------------------------+---------------------------------------------------------------------------------+
| *Independent variable* | -   Leverage                         | -   `hatvalues()` or `fortify()`                                                |
+------------------------+--------------------------------------+---------------------------------------------------------------------------------+
| *Independent variable* | -   Mahalanobis distance             | -   `mahalanobis()`[^regressions-2] or `mahalanobis_distance()`[^regressions-3] |
+------------------------+--------------------------------------+---------------------------------------------------------------------------------+
| *Entire model*         | *Global measures of influence*       | *Global measures of influence*                                                  |
|                        |                                      |                                                                                 |
|                        | -   DFFITS,                          | -   `dffits()`                                                                  |
|                        |                                      |                                                                                 |
|                        | -   Cook's d                         | -   `cooks.distance()` or `fortify()`                                           |
+------------------------+--------------------------------------+---------------------------------------------------------------------------------+
| *Entire model*         | *Specific measures of influence:*    | *Specific measures of influence:*                                               |
|                        |                                      |                                                                                 |
|                        | -   DFBETAS                          | -   `dfbetas()`                                                                 |
+------------------------+--------------------------------------+---------------------------------------------------------------------------------+

: (\#tab:outliers-in-multiple-regressions)Outlier detection in multiple regressions

[^regressions-2]: This function does not take the model as an attribute, but instead requires the data, the column means and the covariance. Thus, we have to specify this function in the following way: `mahalanobis(data, colMeans(data), cov(data))`

[^regressions-3]: Function from `rstatix` package which automatically classifies values as outliers for us.

While it might be clear why we need to use different approaches to find outliers in different components of our model, this might be less clear when evaluating outliers that affect the entire model. We distinguish between *global measures of influence*, which identify how a single observation affects the quality of the entire model, and *specific measures of influence*, which identify how a single observation affects each individual independent variable, i.e. its regression coefficients denoted as $\beta$. It is recommended to look at all different measures of outliers before venturing ahead to performing a linear multiple regression. Going through the entire set of possible outliers would way beyond the scope of this book. So, I will focus on five popular measures which cover all three categories:

-   Dependent variable: Externally studentized residuals

-   Independent variable: Leverage and Mahalanobis distance

-   Entire model: Cook's d, DFBETAS

The approach taken is the same for the other outlier detection methods. Thus, it should be fairly simple to reproduce these as well after having finished the chapters below.

#### Outliers in the dependent variable {#outliers-in-the-dependent-variable}

Irrespective of whether we look at independent or dependent variables, we always want to know whether there are extreme values present. Here I will use the externally studentized residual which 'is the preferred statistic to use to identify cases whose [...] values are highly discrepant from their predicted values [@cohen2014applied, p.401].

First we need to compute the residuals for our model as a new column in our dataset. Since we also want to use other methods to investigate outliers we can use the function `fortify()` which will add some of the later indicators as well and create a `tibble` which only includes the variables from our model. One function to do a lot of things at once. In a second step we hadd the studentized residuals using `rstudent()`.

```{r Compute externally studentized residuals, echo=TRUE}
# Create a tibble with some pre-computed stats
m1_outliers <- fortify(m1) %>% as_tibble()
glimpse(m1_outliers)

# Add the externally studentized residuals
m1_outliers <- m1_outliers %>%
  mutate(.studresid = rstudent(m1))
glimpse(m1_outliers)
```

With our dataset ready for plotting we can do exactly that.

```{r Plot externally studentized residuals, echo=TRUE}
# Create an ID column
m1_outliers <- m1_outliers %>% rownames_to_column()

m1_outliers %>%
  ggplot(aes(x = rowname,
             y = .studresid)) +
  geom_point(size = 0.5)
```

The values of the externally studentized residuals can be positive or negative. All we need to know now is which values count as outliers and which ones do not. @cohen2014applied (p. 401) provides some guidance:

-   general: $outlier = \pm 2$

-   bigger samples: $outlier = \pm 3$ or $\pm 3.5$ or $\pm 4$

As you can tell, it is a matter of well-informed personal judgement. Our dataset consists of over 1200 observations. As such, the dataset certainly counts as large. We can take a look and see how many outliers we would get for each of the benchmarks.

```{r Compute the number of outliers externally studentized residuals, echo=TRUE}
out_detect <- m1_outliers %>%
  mutate(pm_2 = ifelse(abs(.studresid) > 2, "TRUE", "FALSE"),
         pm_3 = ifelse(abs(.studresid) > 3, "TRUE", "FALSE"),
         pm_35 = ifelse(abs(.studresid) > 3.5, "TRUE", "FALSE"),
         pm_4 = ifelse(abs(.studresid) > 4, "TRUE", "FALSE"))

out_detect %>% count(pm_2)
out_detect %>% count(pm_3)
out_detect %>% count(pm_35)
out_detect %>% count(pm_4)
```

The results indicate we could have as many as 53 outliers and as little as 7. It becomes apparent that choosing the right threshold is a tricky undertaking. Let's plot the data again in an easier way to read and add some of the thresholds. I skip `3.5`, since it is very close to `3`. I also reorder the observations (i.e. the x axis) based on `.studresid` using `reorder()`.

```{r Plotting the number of outliers externally studentized residuals, echo=TRUE}
m1_outliers %>%
  ggplot(aes(x = reorder(rowname, .studresid),
             y = .studresid)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = c(-2, 2), col = "green") +
  geom_hline(yintercept = c(-3, 3), col = "orange") +
    geom_hline(yintercept = c(-4, 4), col = "red")
```

At this point, it is a matter of choosing the threshold that you feel is most appropriate. More importantly, though, you have to make sure you are transparent in your choices and provide some explanations around your decision-making. For example, a threshold of `2` appears a bit too harsh for my taste and identifies too many observations as outliers. Using the orange threshold of `3` seems to capture most outliers, because we can also visually see how the dots start to look less like a line and separate out more strongly. Besides, a threshold of `3` seems also more suitable for larger datasets. Since we have an id column (i.e. `rownames`) we can also store our outliers in a separate object to easily reference it later for comparisons with other measures. You will see later why this is handy to have.

```{r Save outliers in a tibble, echo=TRUE}
outliers <- out_detect %>%
  select(rowname, pm_3) %>%
  rename(studresid = pm_3)
```

There are still more diagnostic steps we have to take before we make a final decision on which observations we want to remove or deal with in other ways (see also Chapter \@ref(dealing-with-outliers).

#### Outliers in the independent variables {#outliers-in-the-independent-variables}

To identify outliers in the independent variables we can either use, for example, *Leverage scores*, or use the *Mahalanobis distances*. Both are legitimate approaches and can be computed very easily.

For the leverage scores, we can find them already in our `fortify()`-ed dataset `m1_outliers`. They are in the column `.hat`. An outlier is defined by the distance from the average leverage value, i.e. the further the distance of an observation from this average leverage, the more likely we have to classify it as an outlier. The average leverage is computed as follows:

::: {#average-leverage-equation align="center"}
$average\ leverage = \frac{k + 1}{n}$
:::

In this equation, *k* stands for the number of predictors (i.e. 6) and *n* for the number of observations (i.e. 1188). Therefore, our average leverage can be computed as follows:

```{r Compute average leverage, echo=TRUE}
(avg_lvg <- (6 + 1) / 1188)
```

Also for this indicator we find different approaches to setting cut-off points. While @hoaglin1978hat argue that a distance twice the average counts as an outlier, @stevens2012applied (p.105) suggests that values three times higher than the average leverage will negatively affect the model. The rationale is the same as for the externally studentized residuals: If the thresholds are too low, we might find ourselves with many observations which we would have to further investigate. This might not always be possible or even desirable. However, this should not imply that many outliers are not worth checking. Instead, if there are many, one would have to raise questions about the model itself and whether an important variable needs adding to explain a series of observations that appear to be somewhat 'off'.

Let's plot the leverages and use @stevens2012applied benchmark to draw our reference line.

```{r Plotting leverages, echo=TRUE}
m1_outliers %>%
  ggplot(aes(x = reorder(rowname, .hat),
             y = .hat)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = 3*avg_lvg, col = "orange")
```

As before, we want to know which observations fall beyond the threshold.

```{r Determine observations above average leverage, echo=TRUE}
new_outliers <- m1_outliers %>%
  mutate(avglvg = ifelse(.hat > 3*avg_lvg, "TRUE", "FALSE")) %>%
  select(rowname, avglvg)

# Add new results to our reference list
outliers <- left_join(outliers, new_outliers, by = "rowname")
```

If you look at our reference object `outliers` now, you will notice that there is only one observation that was detected by both methods. Thus, it very much depends on where we look for outliers, i.e. dependent variable or independent variable.

The second method I will cover in this section is the Mahalanobis distance. Luckily the `rstatix` package includes a handy function `mahalanobis_distance()` which automatically detects outliers and presents them to us.

```{r Mahalanobis distance, echo=TRUE}
mhnbs_outliers <- m1_outliers %>%
  select(new_cases:travel) %>%
  rstatix::mahalanobis_distance() %>% 
  rownames_to_column() %>%
  select(rowname, mahal.dist, is.outlier) %>%
  rename(mhnbs = is.outlier)

# Add new results to our reference list
outliers <- left_join(outliers, mhnbs_outliers, by = "rowname")

# We need to remove mahal.dist because it does not indicate
# whether a value is an outlier #data-cleaning
outliers <- outliers %>% select(-mahal.dist)
```

While very convenient, it does pick the cut-off point for us. As we have learned so far, picking the 'right' cut-off point is important. Since the values follow a chi-square distribution, we can determine the cut-off points based on the relevant critical value at the chosen p value. R has a function which allows to find the critical value for our model, i.e. `qchisq()`.

```{r Chi square critical value, echo=TRUE}
(mhnbs_th <- qchisq(p = 0.05,
                   df = 6,
                   lower.tail = FALSE))
```

The `p` value reflects the probability we are willing to accept that our result is significant/not significant. The `df` refers to the degrees of freedom, which in this case relates to the number of independent variables, i.e. 6. Thus, it is fairly simple to identify a cut-off point yourself by choosing the p value you consider most appropriate. The function `mahalanobis_distance()` assumes $p = 0.01$.

If we wanted to plot outliers like we did before, we can reuse our code from above and replace it with the relevant new variables.

```{r Plotting Mahalanobis distance scores, echo=TRUE}
mhnbs_outliers %>%
  ggplot(aes(x = reorder(rowname, mahal.dist),
             y = mahal.dist)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = mhnbs_th, col = "orange")
```

We can notice that the Mahalanobis distance identifies less outliers compared to the leverage, but the same ones. Thus, whether you really need to use both approaches for the same study is questionable and likely redundant. Still, in a few edge cases, you might want to double-check the results, especially when you feel uncertain which observations should be dealt with later.

Besides looking at each side of the regression separately, we might also want to consider whether removing an observation significantly changes the $\beta$ estimates. This is done with outlier detection diagnostics which consider the entire model.

#### Outlier detection considering the entire model: Global measures {#outlier-detection-global-measures}

To assess the global impact of outliers on the entirety of a regression model, Cook's d [@cook1982residuals] is a popular method in the Social Sciences. It measures to which extend a single observation can affect the predictive power of our model to explain all other observations. Obviously, we do not wish to keep observations that make it more difficult to predict most of the other observations correctly. However, as shown in Table \@ref(tab:outliers-in-multiple-regressions), there are different approaches to this, but some are somewhat redundant. For example, @cohen2014applied highlight that DDFITS and Cook's d are 'interchangeable statistics' (p. 404). Thus, there is no point in demonstrating both since they function in a similar way. Your decision might be swayed by the preferences of a publisher, lecturer, supervisor or reviewer. Here, I will focus on Cook's d since it was I see most frequently used in my own field. By all means, feel encouraged to go the extra mile and perform the same steps for the DDFITS.

The good news, `fortify()` automatically added `.cooksd` to our dataset `m1_outliers`. Thus, we cannot only compute whether outliers exist, but also inspect them visually as we did before. A good starting point to find outliers via the Cook's D is to plot its distribution.

```{r Plotting Cooks d, echo=TRUE}
m1_outliers %>%
  ggplot(aes(x = rowname,
             y = .cooksd)) +
  geom_col()
```

Inspecting this barplot, we can tell that there are some observations that have much higher `.cooksd` values than any other observations. Again, we first need to decide on a benchmark to determine whether we can consider these values as outliers. If we follow @cook1982residuals, values that are higher than $d > 1$ require reviewing. Looking at our plot, none of the observations reaches `1` and therefore we would have no need to investigate outliers. Alternatively, @cohen2014applied suggest that other benchmarks are also worth considering, for example based on the critical value of an F distribution, which we can determine with the function `qf()`, which requires us to determine two degrees of freedom ($df_1$ and $df_2$) and a p value. To determine these values @cohen2014applied suggests $p = 0.5$ and the following formulas to determine the correct $df$:

::: {#critical-value-cooks-d align="center"}
$df_1 = k +1$

$df_2 = n - k - 1$
:::

Similar to the critical value for [average leverage](#average-leverage-equation "average leverage"), $k$ reflects the number of predictors, and $n$ refers to the sample size. Thus, we can determine the critical value, and therefore our cut-off point as follows:

```{r Compute the critical value of an F distribution, echo=TRUE}
qf(p = 0.5,
   df1 = 6+1,
   df2 = 1188-6-1)
```

Also this score would imply that we have no particular outliers to consider, because the highest value in `.cooksd` is `r round(max(m1_outliers$.cooksd),2)` compute via \``max(m1_outliers$.cooksd)`\`.

We might be lead to believe that our work is done here, but @cohen2014applied recommends that any larger deviation is worth inspecting and we do notice that relative to other observations there are some cases that appear extreme. Ideally, one would further investigate these cases and compare the results of the regression by removing such extreme cases iteratively. This way one can assess whether the extreme observations truly affect the overall estimates of our model. This would imply repeating steps we already covered earlier when performing multiple regressions with and without outliers. At the end of this chapter of outliers we will rerun the regression to see what effects it had on our model estimates. However, before we can do this, we have one more method of detecting outliers to cover.

#### Outlier detection considering the entire model: Specific measures {#outlier-detection-specific-measures}

While Cook's d helps us to identify outliers which affect the quality of the entire model, there is also a way to investigate how outliers affect specific predictors. This is achieved with DIFBETAS, which also, similar to previous methods, identifies outliers by assessing the impact of deleting extreme cases has on other parts of the regression model.

The function `dbetas()` takes our model `m1` and returns the statistics for each predictor in our model. Thus, we do not receive a single score for each observation but multiple for each specific predictor, i.e. each specific measure.

```{r Computing DFBETAS, echo=TRUE}
m1_dfbetas <- dfbetas(m1) %>%
  as_tibble() %>%               # convert into a tibble for convenience
  rownames_to_column()          # add our rownames

glimpse(m1_dfbetas)
```

All that is left to do is to compare the scores against a benchmark for each variable in this dataset and we know which observations substantially affect one of the predictors. @cohen2014applied provides us with the following recommendations for suitable thresholds:

-   small or moderate dataset: $DFBETAS > \pm 1$

-   large dataset: $DFBETAS > \pm\frac{2}{\sqrt(n)}$

Since our dataset falls rather into the 'large' camp, we should choose the second option. Again, $n$ stands for the sample size. Let's create an object to store this value.

```{r Compute DFBETAS thresholds, echo=TRUE}
(dfbetas_th <- 2 / sqrt(1188))
```

For demonstration purposes, I will pick `masks` to check for outliers. If this was a proper analysis for a project you would have to compute this for each variable separately. As the benchmarks indicate above, the DFEBTAS values can be positive or negative. So, when we compare the computed values with it we need to look at the absolute value, i.e. use `abs()`.

```{r Check for outliers using DFBETAS, echo=TRUE}
# Check whether values exceed the threshold
dfbetas_check <- m1_dfbetas %>%
  mutate(dfbetas_masks = ifelse(abs(masks) > dfbetas_th,
                                "TRUE",
                                "FALSE")) %>%
  select(rowname, dfbetas_masks)

# Add result to our outliers object
outliers <- outliers %>% left_join(dfbetas_check, by = "rowname")

# some housekeeping, i.e. making all columns <lgl> except for rowname
outliers <- outliers %>%
  mutate(rowname = as_factor(rowname)) %>%
  mutate_if(is.character, as.logical)
```

Of course, we can also visualise the outliers as we did before.

```{r Plot outliers using DFBETAS, echo=TRUE}
m1_dfbetas %>%
  ggplot(aes(x = rowname,
             y = masks)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = dfbetas_th, col = "red") +
  geom_hline(yintercept = -dfbetas_th, col = "red")
```

I want to take this opportunity to also show that sometimes we can make visualisations even simpler. Remember, we used `abs()` to make all values positive? We can apply the simple principle here. This way we only need one line to indicate the theshold. In addition, we could also plot multiple variables at once. Instead of defining the `aes()` inside the `ggplot()` function, we can also define it for each `geom` separately. What do you think about the following version of the same plot?

```{r Alternative way of plotting DFBETAS, echo=TRUE}
m1_dfbetas %>%
  ggplot() +
  geom_point(aes(x = reorder(rowname, abs(masks)),
                 y = abs(masks),
                 col = "masks"),
             size = 0.5,
             alpha = 0.5) +
  geom_point(aes(x = rowname,
                 y = abs(travel),
                 col = "travel"),
             size = 0.5,
             alpha = 0.5) +
  geom_hline(yintercept = dfbetas_th, col = "#FF503A") +
  scale_color_manual(values = c("masks" = "#2EA5FF",
                       "travel" = "#7DB33B")) +
  
  # Label the legend appropriately
  labs(col = "COVID measures")
```

After `reorder()`ing the variable `masks` and plotting `travel` as well, we can notice that there are seemingly less outliers for `masks` than for `travel`. Thus, some observations affect some predictors more strongly than others.

#### Reviewing the outliers {#reviewing-the-outliers}

After all this hard work and what turned out to be a very lengthy chapter, we finally arrive at the point to check which observations we might wish to remove or handle in some shape or form. First, we want to know which observations are affected. Therefore we need to review our reference data `outliers`. What we want to see are all observations which were identified by one or more diagnostic tool as an outlier. There is two ways to achieve this. We can use what we have learned so far and simply `filter()` each column for the value `TRUE`. This is the hard way of doing it and if you have many more columns this will take a little while and potentially drive you insane. The easy (and clever) way of filtering across multiple columns can be achieved by turning multiple columns into a single column. In the `tidyverse` this is called `pivot_longer()`. The following steps might sound complicated buy in fact it is relatively simple as long as you remember what 'tidy data' means, i.e. each row represents one observation and each column represents one variable. Let's do this step-by-step. Currently our data has five columns, of which four are different measures to detect outliers. Our goal is to create a `tibble` which has only three columns:

-   `rowname`, which we keep unchanged, because it is the ID which identifies each observation in our data

-   `outlier_measure`, which is the variable that indicates which measure was used to find an outlier.

-   `is.outlier`, which contains the values from the tibble, i.e. the cell values of `TRUE` and `FALSE`.

Here is an example of what we want to achieve. Imagine we have a very small dataset, which contains three columns, two of which are outlier detection measures, i.e. `studresid` and `mhbns`:

```{r Create example tibble for pivoting, echo=FALSE}
data <- tribble(
  ~rowname,   ~ studresid,    ~mhbns,
         "1",       FALSE,     FALSE,
         "2",        TRUE,     FALSE,
         "3",        TRUE,      TRUE
)
```

```{r Outlier data prior to pivoting, echo=TRUE}
data
```

We now want to turn this dataset into a dataset with two columns. One column captures the `rowname` and the other one the outlier detection method. Thus, for every row in this dataset we have two values recorded. For example, the first observation has the values `studresid == FALSE` and `mhnbs == FALSE`. If we wanted to combine these two observations, we need to have two rows which have the `rowname` `1` to ensure that each row is still only contains one observation, i.e. a tidy dataset. Therefore, we end up with more rows than in our original dataset, hence, a 'longer' dataset. Here is how we can do this automatically:

```{r outlier data after pivoting, echo=TRUE}
data <- data %>%
  pivot_longer(cols = !rowname,
               names_to = "outlier_measure",
               values_to = "is.outlier")

data
```

Instead of three rows we have six and all the outlier detection values (i.e. all `TRUE` and `FALSE` values) are now in one column. However, what exactly did just happen in this line of code? In light of what we specified above we did three things inside the function `pivot_longer()`:

-   We excluded `rownames` from being pivoted, i.e. `cols = !rowname`.

-   We specified a column where all the column names go to, i.e. `names_to = "outlier_measure"`.

-   We specified a column where all cell values should be listed, i.e. `values_to = "is.outlier`.

From here, it is very easy to just `count()` the number of `is.outlier` per `rowname` and `filter()` out those that return a value of `TRUE`.

```{r count outliers after pivoting, echo=TRUE}
data %>%
  count(rowname, is.outlier) %>%
  filter(is.outlier == "TRUE")
```

Et voil. We now counted the number of times an observation was detected by all the different measures we use. This is obviously scaleable to more than two measures. Thus, we can apply it to our data as well.

```{r Reviewing the outliers, echo=TRUE}
outliers_true <- outliers %>%
  pivot_longer(!rowname, names_to = "measure", values_to = "is.outlier") %>%
  count(rowname, is.outlier) %>%
  filter(is.outlier == "TRUE")

outliers_true
```

The result is a `tibble` which tells us that there we 72 distinct outliers identified. This might sound like a lot, but we also have to remember that our dataset consists of 1188 observations, i.e. 6% of our data are outliers.

Since the output is fairly long, we might want to plot the outcome to get an idea of the bigger picture.

```{r Plot number of outliers, echo=TRUE}
outliers_true %>%
  ggplot(aes(x = reorder(rowname, n),
             y = n)) +
  geom_col() +
  coord_flip()
```

There are a few observations which were detected by three out of the four methods we used to detect outliers. there are some more which were detected by two different methods. However, there are also more than half of observations which were only detected by one method.

As highlighted in Chapter \@ref(dealing-with-outliers) there are many ways we can go about outliers. To keep our analysis simple, I intend to remove those observations which were detected by multiple methods rather than only by one. Whether this approach is truly appropriate is a matter of further investigation. I take a very pragmatic approach with our sample data. However, it is worth reviewing the options available to transform data, as covered by @field2013discovering (p.203). Data transformation can help to deal with outliers, instead of removing or imputing them. However, data transformation comes with severe drawbacks and in most cases, using a *bootstrapping* technique to account for violations of parametric conditions is often more advisable. Bootstrapping refers to the process of randomly resampling data from your dataset and rerun the same test multiple times, e.g. 2000 times. The process is somewhat similar to multiple imputation (see Chapter \@ref(replacing-removing-missing-data), but instead of estimating a specific value in our dataset, we estimate statistical parameters, e.g. confidence intervals or regression coefficients (i.e. estimates). The package `rsample` allows to implement boostrapping in a fairly straightfoward way. [PROVIDE AN EXAMPLE AS A CASE STUDY AT THE END?]

For now, I settle on removing outliers entirely. The easiest way to remove these outliers is to combine our `outliers` dataset with the one for our regression (`m1_outliers`).

```{r Remove outliers from dataset for multiple regression, echo=TRUE}
# Include outliers which were detected by multiple methods
outliers_select <- outliers_true %>%
  filter(n > 1)

# Keep only columns which are NOT included in outliers_select
m1_no_outliers <- anti_join(m1_outliers, outliers_select, by = "rowname")

m1_no_outliers
```

The function `anti_join()` does exactly what the name implies. It takes the first data frame (i.e. `m1_outliers`) and removes values that are included in the second data frame (i.e. `outliers_select()`). This is tremendously helpful when having performed such complex outlier detection and want to remove them all in one go.

As the final step, we want to compare how the removal of outliers affected our model. Ideally, we managed to improve the regression.

```{r Comparing the multiple regression before and after outlier removal, echo=TRUE, warning=FALSE}
# Original regression
m1 <- lm(new_cases ~ masks + movements + gatherings + schools + businesses + travel,
         data = covid_uk_ger)

# Regression with outliers removed
m2 <- lm(new_cases ~ masks + movements + gatherings + schools + businesses + travel,
         data = m1_no_outliers)

# Compare the parameters between models
parameters::compare_parameters(m1, m2) %>%
  as_tibble() %>%
  select(Parameter, Coefficient.m1, Coefficient.m2)

# Compare the performance between models
performance::compare_performance(m1, m2)
```

If you look at `R2 (adj.)` you might feel a sense of disappointment. How does the model explain less variance? Well, if we consider `AIC`, `BIC` and `RMSE`, we managed to improve the model, because their values are lower for `m2`. However, it seems that the outliers affected `R2`, which means they inflated this model indicator. In other words, our regression explains less than we hoped for. However, we can be more confident that the predictions of this model will be more accurate.

As a final remark, it is worth noting that the $\beta$ for some predictors has changed drastically, e.g. `masks` and `travel`. Thus, these are now more accurate as well and would require a reinterpretation in terms of which predictors have a strong impact on the outcome, i.e. `new_cases`.

### Standardised $\beta$ vs. unstandardised $\beta$

In multiple regressions we often use variables which are measured in different ways and which use different measurement units, e.g. currency, age, etc. Thus, it is fairly difficult to compare their $\beta$s without standardising the measurement units. Standardised scores (also referred to as *z-scores*) for a variable, imply that each variable has a distribution where the mean equals 0 and the standard deviation equals 1. By transforming your variables to z-scores makes them 'unit free' [@cohen2014applied, p.25] and therefore we can easily compare them irrespective of their original measurement unites. Here is an example with a simplified dataset:

```{r Standardisation of a variable z scores, echo=TRUE}
# Create some data
data <- tibble(x = c(1,2,3,4))

# Compute z-scores
data <- data %>%
  mutate(x_scaled = scale(x))

# Compare the mean and sd for each variable
# and put it into a nice table

tibble(
  variable = c("x", "x_scaled"),
  mean = c(mean(data$x), mean(data$x_scaled)),
  sd = c(sd(data$x), sd(data$x_scaled))
)
```

If you feel that my choice of values has likely affected the outcome, feel free to change the `x` values to whatever you like. The results for `x_scaled` will always remain the same. There is something else that remains the same: the actual distribution. Because we are only rescaling our data, we are obviously not affecting the differences between these scores. We can show this in a scatterplot and by running a correlation correlation. Both will show that both variables are perfectly correlated with each other, which means our transformations had not effect on the relative relationship of values to each other.

```{r Similarities of scaled and unscaled data, echo=TRUE}
data %>%
  ggplot(aes(x = x,
             y = x_scaled)) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE)

correlation::correlation(data)
```

For regressions, it is not a bad idea to report both, unstandardised ($B$) and standardised ($\beta$) values for each predictor. However, one could argue that it is a bit redundant. If you only want to report one estimate, it is common to provide the $\beta$ values rather than the unstandardised equivalent.

Unfortunately, this is not the end of the story, because 'standardisation' can mean different things. We can generate standardised $\beta$ after the fact (post-hoc), or decide to 'refit' our regression with standardised predictors, i.e. before we run the regression. Both options are appropriate, but the scores will differ slightly. However, the main idea remains the same: We want to be able to compare different regression coefficients with each other. In R there are many different packages which offer standardised estimates. On package I particularly recommend is `parameters`. It offers different options of how to return our estimates from a linear model by specifying the `standardize` argument (see [detailed documentation](https://easystats.github.io/parameters/reference/model_parameters.default.html "detailed documentation"){target="blank"}). Of course, you can also opt to scale the variables by hand, using the `scale()` function, like we did in the previous example. Both approaches are shown in the following code chunk.

```{r Retrieving standardised betas, echo=TRUE}
# Scaling 'by hand'
m2_scaled <- lm(scale(new_cases) ~
                  scale(masks) +
                  scale(movements) +
                  scale(gatherings) +
                  scale(schools) +
                  scale(businesses) +
                  scale(travel),
                data = m1_no_outliers)

parameters::model_parameters(m2_scaled)

# Scaling using 'parameters' package
parameters::model_parameters(m2, standardize = "refit")

# Scaling using `parameters` package without refitting the model
parameters::model_parameters(m2, standardize = "posthoc")

```

Personally, I prefer to standardise using the `parameters` package, because it makes my life easier and I need to write less code. I can also easily compare results prior to scaling my variables. The choice is yours, of course.

## Multicollinearity: The dilemma of highly correlated independent variables

We finished fitting our model and we are likely exhausted, but also happy that we accomplished something. However, we are not yet done with our analysis. One of the most important post-tests for multiple regressions is a test for multicollinearity or sometimes referred to as collinearity. The phenomenon of multicollinearity defines a situation in which some of our independent variables can be explained by other independent variables in our regression model. In other words, there exists a very large correlation (a linear relationship) between two or more independent variables. It is important to not confuse this with correlations between dependent variable and independent variables. Remember, a regression reflects the linear relationship between the predictor variables and the outcome variable. As such, we do hope to find a correlation between these variables, just not among the independent variables.

If we have evidence that multicollinearity exists in our data, we face some problems:

-   We cannot trust our regression coefficients, i.e. our estimates ($\beta$).

-   Since the same variance can be explained by two different independent variables, it is unclear which one is important, since both can be easily interchanged without affecting the model all that much.

-   It is impossible to test any hypotheses, since we cannot trust our estimates.

-   We also likely underestimate the variance our model can explain, i.e. our $R^2$.

-   We produce more Type II errors, i.e. we likely reject important predictors due to statistical insignificance.

The Variance Inflation Factor (VIF) iand Tolerance are methods to identify issues of multicollinearity. Compared to detecting outliers, it is very simple to compute these indicators and surprisingly uncomplicated. The package `performance` offers a simple function called `check_collinearity()` which provides both. The tolerance is computes as $\frac{1}{VIF}$.

```{r Detecting multicollinearity using VIF, echo=TRUE}
performance::check_collinearity(m2)
```

As the output already indicates, there is a `Low Correlation` between our independent variables. So, good news for us. In terms of interpretations we find the following benchmarks as recommendations to determine multicollinearity:

-   `VIF > 10`: Evidence for multicollinearity

-   `mean(VIF) > 1`: If the mean of all VIFs lies substantially above `1`, multicollinearity might be an issue.

-   `Tolerance < 0.1`: Multicollinearity is a serious concern and is the same condition as `VIF > 10`.

-   `Tolerance < 0.2`: Multicollinearity could be a concern.

However, @cohen2014applied warns to not rely on these indicators alone. There is sufficient evidence that lower VIF scores could also cause issues due to multicollinearity. Thus, it is always important to still investigate relationships of independent variables statistically (e.g. correlation) and visually (e.g. scatterplots). Outliers often can be a cause for too high or too low VIFs. In short, the simplicity in computation in R of these indicators should not be mistaken as a convenient shortcut.

Besides the VIF and tolerance, multicollinearity can also exist among residuals, i.e. the error terms that are associated with each predictor. Regressions assume that errors are independent from each other. The Durbin-Watson Test offers an equally easy way to detect multicollinearity among residuals. We can use the `car` package and the function `durbinWatsonTest()` to retrieve the relevant information.

```{r Durbin Watson Statistics, echo=TRUE}
car::durbinWatsonTest(m2) %>%
  broom::tidy()
```

The `statistic` of the Durbin-Watson Test can range from 0 to 4, where 2 indicates no correlation. Luckily for us, we do not have to guess whether the difference from our computation is significantly different from 2, because we also get the `p.value`. In our case, we find that our model suffers from multicollinearity of error terms.

If we wanted to remedy multicollinearity we could consider for example:

-   Revisiting the regression model and potentially dropping those variables we know measure the same or similar underlying factors [@cohen2014applied].

-   We could collect more data, because a larger dataset will always increase the precision of the regression coefficients [@cohen2014applied].

-   Use different modelling techniques which can resolve such issues, for example using generalised least squares (GLS) and transform some variables [@sage-methods2004dbw].

The cause for this multicollinearity of residuals can be rooted in either 'clustered data' or a 'serial dependency' [@cohen2014applied]. In our dataset, both could apply. First, we have data that was collected over time which could lead to wrong standard errors. Second, we have data from two different countries (`United Kingdom` and `Germany`). Thus, our data might be clustered.

In order to remove serial depenceny effects we would have to transform data and account of the correlation across time. Such a technique is showen in @cohen2014applied (p.149). To counteract the issue of clustered data, we need to use multilevel regression models, also known as hierarchical regressions, which we will cover in the next chapter.

## Hierarchical regression

```{r Hierarchical regression, echo=TRUE}

```

## Moderated regression {#moderated-regressions}

```{r Moderated regression, echo=TRUE}

```
