# Regression: Creating models to predict future observations {#regression}

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(r4np)
```

Regressions are an interesting area of data analysis since it enables us to make very specific predictions about the future incorporating different variables at the same time. As the name implies, regressions 'regress', i.e. they draw on past observations to make predictions about future observations. Thus, any analysis incorporating a regression makes the implicit assumption that the future can be best explained by the past.

I once heard someone refer to regressions as driving a car by looking at the rearview mirror. As long as the road is straight, we will be able to successfully navigate the car. However, if there is a sudden turn, we might drive into the abyss. This makes it very clear when and how regressions are can be helpful.

Regressions are also a machine learning method, which falls under models with supervised learning. If you find machine learning fascinating, you might find the book "Hands-on Machine Learning with R" [@boehmke2019hands] very insightful and interesting.

In the following chapters we will cover three common types of regressions:

-   Single linear regression

-   Multiple regression

-   Hierarchical regression

These three types will allow you to perform any other type of linear regression you could think of. We can further distinguish two approaches to modelling via regressions:

-   *hypothesis testing*: A regression model is defined ex-ante

-   *machine learning*: A model is developed based on empirical data

In the following chapters we will slightly blur the lines between both approaches.

All our regressions will be performed using the `covid` dataset of the `r4np` package to investigate whether certain factors can predict COVID numbers in different countries. I felt, this book would not have been complete without covering this topic. After all, I wrote this book during the pandemic and it likely will mark a dark chapter in human history.

## Single linear regression {#single-linear-regression}

A single linear regression looks very similar to a correlation (see Chapter \@ref(correlations), but it is different in that it defines which variable affects another variable, i.e. a single direction relationship. I used the terms dependent variable (DV) and independent variable (IV) previously when comparing groups (see Chapter \@ref(comparing-groups), and we will use them here again. In group comparisons, the independent variable was usually a `factor`, but in regressions we can use data that is not a categorical variable, i.e. `integer`, `double`, etc.

While I understand that mathematical equations can be confusing, with regressions, they are fairly simple to understand. Also, when writing our models in R, we will continuously use a `formula` to specify our regression. A single linear regression consists of one independent variable and one dependent variable:

::: {align="center"}
$$
DV = \beta_{0} + IV * \beta_{1} + error
$$
:::

Beta ($\beta$) represents the coefficient of the independent variable, i.e. how much a change in IV causes a change in DV. For example, a one unit change in the IV might mean that the DV changes by two units of IV:

::: {#single-linear-regression-example align="center"}
$$
DV = \beta_0 + 2 * IV + error
$$
:::

If we ignore $\beta_0$ and $error$ for a moment, we find that that if $IV = 3$, our $DV = 2*3 = 6$. Therefore, if $IV = 5$, we find that $DV = 10$, and so on. According to this model, DV will always be twice as large as IV.

You might be wondering what $\beta_0$ stands for. It indicates an offset for each value, also called the intercept. Thus, no matter which value we choose for IV, DV will always be $\beta_0$ different from IV. It is a constant in our model. This can be best explained by visualising a regression line. Pay particular attention to the to the expressions after `function(x)`

```{r beta zero and beta one explained, echo=TRUE}
# A: Two models with different beta(0)
ggplot() +
  geom_function(fun = function(x) 2 * x, colour = "red") +
  geom_function(fun = function(x) 1 + 2 * x, colour = "blue") +
  see::theme_modern()

# B: Two models with the same beta(0), but different beta(1)
ggplot() +
  geom_function(fun = function(x) 2 * x, colour = "red") +
  geom_function(fun = function(x) 3 * x, colour = "blue") +
  see::theme_modern()
```

Plot B also shows what happens if we change $\beta_1$, i.e. the slope. The two models both have the same intercept (and therefore the same origin), but the blue line ascends quicker than the red one, because its $\beta_1$ is higher than the one for the red model.

Lastly, the $error$ component in the regression model refers to the deviation of data from this regression lines. Ideally, we want this value to be as small as possible.

### Fitting a regression model by hand, i.e. trial and error

If this sounds all awfully theoretical, let's try to fit a regression model by hand. First we need to consider what our model should be able to predict. Let's say that the number of COVID-19 cases predicts the number of deaths due to COVID-19. Intuitively we would assume this should be a linear relationship, because the more cases there are, the more likely we find more deaths caused by it.

```{r Finding a regression equation by hand p1, echo=TRUE}
# We only select most recent numbers, i.e. "2021-08-26"
# and countries which have COVID cases
covid_sample <- covid %>%
  filter(date_reported == "2021-08-26" &
           cumulative_cases != 0)

covid_sample %>%
  ggplot(aes(x = cumulative_cases,
             y = cumulative_deaths)) +
  geom_point()
```

This data visualisation shows us not much. We can see that there are three countries, which appear to have considerably more cases than most other countries. Thus, all other countries are crammed together in the bottom left corner. To improve this visualisation without removing the outliers, we can rescale the x and y axis using the function `scale_x_continuous()` and `scale_y_continuous()`.

```{r Finding a regression equation by hand p2, echo=TRUE, warning=FALSE}
covid_sample %>%
  ggplot(aes(x = cumulative_cases,
             y = cumulative_deaths)) +
  geom_point() +
  scale_x_continuous(trans = "log") +
  scale_y_continuous(trans = "log")
```

As we can see, the scatterplot is now easier to read and the dots are more spread out. This reveals that there is quite a strong relationship between `cumulative_cases` and `cumulative_deaths`. However, similar to before, we should avoid outliers when performing our analysis. For the sake of simplicity, in this section I will limit the number of countries included in our analysis, which also removes the requirement of using `scale_x_continuous()` and `scale_y_continuous()`.

```{r Finding a regression equation by hand p3, echo=TRUE, warning=FALSE}
covid_sample <- covid_sample %>%
  filter(date_reported == "2021-08-26" &
           cumulative_cases >= 2500 &
           cumulative_cases <= 150000 &
           cumulative_deaths <= 3000)

plot <- covid_sample %>%
  ggplot(aes(x = cumulative_cases,
             y = cumulative_deaths)) +
  geom_point()

plot
```

Through trial and error we can try to fit a linear line on top by adjusting the beta values. This is effectively what we hope to achieve with a regression: the best $\beta$ values which best explain our data. Let's start with the basic assumption of $y = x$ without specific $\beta$s, i.e. they are zero.

```{r Finding a regression equation by hand p4, echo=TRUE, warning=FALSE}
plot +
  geom_function(fun = function(x) x, colour = "red")
```

What we try to achieve is that the red line fits nicely inside the cloud of dots. Our very simple model provides a very poor fit to our data points, because all the dots are way below it. This makes sense, because $y = x$ would imply that every COVID-19 case leads to a death, i.e. everyone with COVID did not survive. From our own experience we know that this is luckily not true. Ideally we want the line to be less steep, because our first model does not make much sense. We can do this by adding a $\beta_1$ to our equation. Maybe only 2% of people who got COVID-19 might not have recovered, i.e. $\beta_1 = 0.02$.

```{r Finding a regression equation by hand p5, echo=TRUE, warning=FALSE}
plot +
  geom_function(fun = function(x) 0.02*x, colour = "red")
```

This time the line looks much more aligned with our observations. One could argue that it might have to move a little to the right as well, to cover the observations at the bottom a bit better. Therefore, we should add a $\beta_0$ to our equation, e.g. `-50`.= to move it to the right and tweak the $\beta_1$ ever so slightly.

```{r Finding a regression equation by hand p6, echo=TRUE, warning=FALSE}
plot +
  geom_function(fun = function(x) -50 + 0.015*x, colour = "red")
```

We finished creating our regression model. If we wanted to express it as a formula we would write $DV = -5 + 0.015 * IV$. We could now use this model to predict how high COVID cases likely will be in other countries.

Estimating a regression model in this way is not ideal and far from accurate. Instead, we would compute the $\beta$s based on our observed data, i.e. `cumulative_cases` and `cumulative_deaths`. We can use the function `lm()` to achieve this. I also rounded the all `numeric` values to two decimal places to make the output easier to read. We also use `tidy()` to retrieve a cleaner output from the computation.

```{r Computing a regression, echo=TRUE}
# classic r
m0 <- lm(cumulative_deaths ~ cumulative_cases, data = covid_sample)
broom::tidy(m0) %>% mutate(across(where(is.numeric), round, 2))
```

We first might notice that the `p.value` indicates that the relationship between `cumulative death` and `cumulative_cases` is significant. Thus, we can conclude that countries with more COVID cases also suffer higher numbers of people who do not successfully recover from it. However, you might be wondering where our $\beta$ scores are. They are found where it says `estimate`. The standard error (`std.error`) denotes the `error` we specified in the [previous equation](#single-linear-regression). We find that in the first row we get the $\beta_0$, i.e. the one for the intercept which is 88.10. This one is larger than what we estimated, i.e. `-50`. However, $\beta_1$ is 0.01, which means we have done a very good job in guessing this estimate. Still, it becomes hopefully obvious that it is much easier to use the function `lm()` to estimate a model than 'eyeballing it'.

We can now visualise the computed model (in blue) and our guessed model (in red) in one plot and see the differences. The plot shows that we have not been too far off. However, it was relatively easy to fit a model onto the observed data in this case. Often, it is much more difficult, especially when more than two variables are involved.

```{r Showing computed and guessed regression, echo=TRUE}
plot +
  geom_function(fun = function(x) -50 + 0.015*x, colour = "red") +
    geom_function(fun = function(x) 88.1 + 0.01*x, colour = "blue")
```

With our final model computed, we also need to check its quality in terms of predictive power based on how well it can actually explain our observed data. We have tested models before when we looked at confirmatory factor analyses for latent variables (see Chapter \@ref(latent-constructs). This time we want to know how accurate our model is in explaining observed data and therefore how accurate it will be predicting future observations. The package `performance` offers a nice little shortcut to compute many different things at once:

-   `check_model()`: Checks for linearity, homogeneity, collinearity and outliers

-   `model_performance()`: Tests the quality of our model.

For now, we are mainly interested in the performance of our model. So, we can compute it the following way:

```{r Single regression model performance, echo=TRUE}
performance::model_performance(m0)
```

There are quite a number of performance indicators and here is how to read them:

-   `AIC` stands for *Akaike Information Criterion* and the lower the score the better the model

-   `BIC` stands for *Bayesian Information Criterion* and the lower the score the better the model

-   `R2` stands for *R squared* ($R^2$) and is also known as the coefficient of determination. It measures how much the independent variable can explain the variance in the dependent variable. In other words, the higher $R^2$ the better is our model, because more of the variance can be explained by our model. $R^2$ falls between 0-1, where 1 would imply that our model can explain 100% of the variance in our sample. $R^2$ is also considered a *goodness-of-fit measure*.

-   `R2 (adj.)` stands for *adjusted R squared*. The adjusted version of $R^2$ becomes important if we have more than one preditor (i.e. independent variable) in our regression. The adjustment of $R^2$ accounts for the number of independent variables in our model. Thus, it is possible for us to compare different models with each other, even though they might have different numbers of predictors. It is important to note that $R^2$ will always increase if we add more predictors.

-   `RMSE` stands for *Root Mean Square Error* and is an indicate how small or large the prediction error of the model is. Conceptually, it aims to measure the average deviations of values from our model when we attempt predictions. The lower the score the better, i.e. a score of 0 would imply that our model perfectly fits the data, which is likely never the case in the field of Social Sciences. The RMSE is particularly useful when trying to compare models.

-   `Sigma` stands for the standard deviation of our residuals (the difference between predicted and empirically observed values) and is therefore a measure of prediction accuracy. `Sigma` is *'a measure of the average distance each observation falls from its prediction from the model'* [@gelman2020regression, p. 168].

Many of these indices will become more relevant when we compare models. However, $R^2$ can also be meaningfully interpreted without a reference model. We know that the bigger $R^2$ the better. In our case it is `0.548` which is very good considering that our model consist of only one predictor only. It is not easy to interpret whether an particular $R^2$ value is good or bad. In our simple single linear regression, $R^2$ is literally 'r squared', which we already know from correlations and their effect sizes (see Table \@ref(tab:effect-size-cohen). Thus, if we take the square root of $R^2$ we can retrieve the correlation coefficient, i.e. $r = \sqrt{R^2} = \sqrt{0.548} = 0.740$. According to @cohen1988statistical, this would count as a large effect size.

However, for multiple regressions, the situation is slightly more complicated, but the interpretation of $R^2$ and its adjusted version remain largely the same.

Once you have a model and it is fairly accurate, you can start making predictions. This can be achieved by using our model object `m0` in combination with the function `predict()`. However, first we should define a set of values for our independent variable, i.e. `cumulative_cases`, which we store in a tibble using the `tribble()` function.

```{r Predict values from linear regression, echo=TRUE}
df_predict <- tribble(
  ~cumulative_cases,
                100,
               1000,
              10000,
             100000
)
predict(m0, newdata = df_predict)
```

As a result we find out how many likely deaths from COVID have to be expected based on our model for each value in our dataset.

Single linear regressions are simple and a good way to introduce novice scholars to modeling social phenomena. However, hardly ever will find that a single variable can explain enough variance to be a useful model. Instead, we most likely can improve the majority of our single regression models by considering more variables in the form of multiple regressions.

## Multiple regression

Multiple regressions expand single linear regressions by allowing us to add more variables. Maybe less surprising, computing a multiple regression is similar to a single regression in R, because it requires the same function, i.e. `lm()`. However, we add more IVs. Therefore, the equation we used before needs to be modified slightly by adding more independent variables each of which will have its own $\beta$ value:

::: {align="center"}
$$
DV = \beta_{0} + IV_{1} * \beta_{1} + IV_{2} * \beta_{2} + ... + IV_{n} * \beta_{n} + error
$$
:::

In the last section we wanted to know how many people will likely not recover from COVID. However, it might be even more interesting to understand how we can predict new cases and prevent casualties from the outset. Since I live in the United Kingdom at the time of the pandemic, I am curious to know whether certain COVID measures helped to reduce the number of new cases there. To keep it more interesting, I will also add Germany to the mix since it has shown to be very effective in handling the pandemic relative to other European countries. Of course, feel free to pick different countries (maybe the one you live in?) to follow with my example. In Chapter \@ref(moderated-regressions) it will become obvious why I chose two countries (\#spoiler-alert).

First, we create a dataset that only contains information from the `United Kingdom` and `Germany`[^regressions-1], which means we use `filter()`.

[^regressions-1]: Germany is called 'Deutschland' in German, hence the abbreivation 'DEU' according to the [ISO country code](https://www.iso.org/obp/ui/#iso:code:3166:DE "ISO country code"){target="blank"}.

```{r MR filter out UK and DEU as a sample}
covid_uk_ger <- covid %>%
  filter(iso3 == "GBR" | iso3 == "DEU")
```

In a next step, we might want to know how new cases are distributed over time. Let's plot this first.

```{r Multiple regression plotting new cases, echo=TRUE}
covid_uk_ger %>%
  ggplot(aes(x = date_reported,
             y = new_cases)) +
  geom_col()
```

We can tell that there are different waves of new cases. As such, we should find that there are certain variables that should help explain when `new_cases` are high and when they are low. If you have hypotheses you want to test, you would already know which variables to include in your regression. However, in our case, we do not really have a hypothesis based on our prior reading or other studies. Thus, we pick variables of interest that we suspect could help us with modelling new COVID cases. We can be fairly certain that the number of new COVID cases should be lower if there are more safety measures in place - assuming that they are effective, of course. The `covid` dataset includes such information evaluated by the [WHO](https://covid19.who.int/info/ "WHO"){target="blank"}, i.e. `masks`, `travel`, `gatherings`, `schools` and `movements`. Remember, you can always find out what these variables stand for by typing `?covid` into the console.

We simply add these variables by using the `+` symbol in the `lm()` function.

```{r Multiple regression, echo=TRUE}
# Create our model
m1 <- lm(new_cases ~ masks + movements + gatherings + schools + businesses + travel,
         data = covid_uk_ger)

# Inspect the model specifications
broom::tidy(m1)

# Evaluate the quality of our model
performance::model_performance(m1)

# covid_uk_ger %>% correlation::correlation(redundant = TRUE)
```

Overall, the model is not particularly great, because even though we added so many variables, the $adjusted \ R^2$ is not particularly high, i.e. only 0.229. There are a couple more important insights gained from this analysis:

-   All variables appear to be significant and therefore help to explain new cases.

-   The variables `masks`, `movements` and `gatherings` seem to reduce `new_cases`, i.e. they have negative estimates ($\beta$).

-   However, `schools`, `businesses`, `travel` have a positive effect on `new_cases`.

Especially the last point might appear confusing. How can measures taken increase the number of new COVID cases? Should we avoid them? What we have not considered in our regression is the fact that measures might be put in place to reduce the number of new cases rather than the other way around. Thus, it might not be that `schools`, `business` and `travel` predict higher `new_cases`, but rather the opposite, i.e. due to higher `new_cases` the measures for `schools`, `businesses` and `travel` were tightened, which later on (with a time lag) led to lower `new_cases`. Thus, the relationships might be a bit more complicated, but to keep things simple, we accept the fact that with our data we face certain limitations (as is usually the case).

However, there are a couple of things we overlooked when running this regression. We did not test for sources of bias, especially outliers. Since multiple regressions consider more than one variable we have to consider outliers in light of other variables. As such, we have to draw on different tools to assess our data. There are certain pre- and post-tests we have to perform to fully assess a multiple regression model:

-   pre-test: We need to consider whether there are any outliers and whether all variables are parametric

-   post-test: Do our independent variables correlate very strongly with each other, i.e. are there issues of multiple collinearity.

We already covered aspects of linearity, normality and homogeneity of variance. However, outliers and collinearity have to be reconsidered for multiple regressions.

### Outliers in multiple regressions

While it should be fairly clear by now why we need to handle outliers (remember Chapter \@ref(dealing-with-outliers)), the approach we take is somewhat different when we need to consider multiple variables at once. Instead of identifying outliers for each variable independently, we have to consider the interplay of variables as well. In short, we need a different technique to assess outliers. For regressions Cook's d is the preferred method in the Social Sciences. Our plotting package `ggplot2` has a nice function to automatically add model fit statistics to your dataset based on our regression model, i.e. `fortify()`.

```{r Computing Cooks d, echo=TRUE}
m1_fitted <- fortify(m1) %>% as_tibble()
glimpse(m1_fitted)
```

Among other variables, `fortify()` automatically adds `.cooksd`. Thus, we cannot only compute whether outliers exist, but also inspect them visually as we did before. A good starting point to find outliers via the Cook's D is to plot the distribution.

```{r Plotting Cooks d, echo=TRUE}
m1_fitted %>%
  ggplot(aes(x = rownames(m1_fitted),
             y = .cooksd)) +
  geom_col()
```

In order to plot every observation in our dataset, we had to use the function `rownames()`, which in our case represents a consecutive number which indexes each row, like an id column in a dataset. We use this 'row number' for our x axis. Here is what `rownames(m1_fitted)` produces:

```{r Show what rownames does, echo=TRUE}
rownames(m1_fitted)
```

If you want this sequence as a regular column in your dataset, you can use the handy function `rownames_to_column()` from the `tibble` package, which is part of the `tidyverse`.

Inspecting this barplot, we can tell that there are some observations that are far away from any other values. However, we first need to decide on a benchmark to determine whether we can consider these values as outliers.

## Hierarchical regression

```{r Hierarchical regression, echo=TRUE}

```

## Moderated regression {#moderated-regressions}

```{r Moderated regression, echo=TRUE}

```
